---
title: "Snowshoe Exploratory Report"
author: "Emiliano Espinoza"
date: "11/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message =FALSE, warning = FALSE)
library(tidyverse)
library(janitor)
library(here)
library(ggpubr)
library(broom)
library(effsize)
library(broom)
library(palmerpenguins)
library(ggbeeswarm)
library(kableExtra)
library(lubridate)
library(scales)

snowshoe <- read_csv(here("data", "bonanza_hares.csv"))
```

### 1. Overview

Snowshoe hares, Lepus americanus, are a 'keystone' prey species in northern boreal forests and experience population fluctuations of 8-11-years. Despite intense responses of both vegetation and predators to changes in hare densities, landscape-scale comparisons of hare populations in Alaska have been limited to qualitative descriptions. This study was unable to detect declines in apparent survival during declining densities in study populations.This report provides an exploratory overview of differences in juvenile snowshoe hare sizes observed in 5 locations in the Tanana valley, Alaska from 1999 to 2012, via capture-recapture method. (Kielland et al. (2017)). Specifically, the relationship(s) between hind foot lengths and body weight were explored.

### 2. Data and Analyses

Juvenile Snowshoe Hares size and weight measurements were collected and made available by Dr. Kielland Knut at 5 locales in the Tanana valley located within the Bonanza Creek Experimental Forest from 1998 to 2012. The data contain observations for 378 juvenile snowshoe hares, collected from 199-2012, excluding years 2002 and 2009. 
Following exploratory data visualization, body weight distributions between male and female juvenile hares are explored by site, with summary statistics accompanying visualizations. The relationship between hind foot lengths and body weights are explored by simple linear regression. All analyses are in R version 4.0.2 using RStudio version 1.3.1056. 

### 3. Exploratory Findings

#### A. Juvenile Snowshoe Hare Trap Counts

```{r}
juvenile <- snowshoe %>% #Filtering data to only display juvenile ages
  filter(age == "j") %>% 
  mutate(date = mdy(date)) %>% #Doesn't like YMD format
  mutate(year = year(date))

juvenile_count <- juvenile %>% #Gathering count summary statistics of juvenile observations.
  count(year) %>% 
  summarize(
    mean_count = mean(n),
    median_count = median(n),
    sd_count = sd(n),
    max_count = max(n),
    min_count = min(n))
```

**Table 1.** Descriptive statistics (mean, standard deviation, and sample size) of annual juvenile hare trappings per year. Data: Kielland et al. (2017).

```{r descriptive_stats}
juvenile_count %>% #Creating table that conveys summary statistics gathered.
  kable(col.names = c("Mean count per year", "Median count per year", "Standard deviation", "Maximum per year", "Minimum per year")) %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r juvenile_count_plot, fig.align = "center"}
ggplot(data = juvenile, aes(x = year)) + #Creating visualization of hare trap counts by year
  geom_histogram(fill = "blue") +
  theme_minimal() +
  labs(x = "Year",
       y = "Juvenile Trappings",
       title = "Annual Juvenile Hare Trap Counts") +
  scale_x_continuous(breaks = pretty_breaks(n = 12)) + # Allows you to further depict information of x axis.
  scale_y_continuous(breaks = pretty_breaks(n = 7)) # ""
```

**Figure 1.** Juvenile Snowshoe Hare trapping count (y-axis) per year of study (x-axis), excluding years 2002 and 2009 in which there is no data collected for those years.
Data: Kielland et al. (2017). 

Table 1 and Figure 2 depict trapping count data from all sites across the years collected for the study. The varied distributions conveyed minimum (`r juvenile_count$min_count`, in 2010) and maximum (`r juvenile_count$max_count`, in 1999) counts of juvenile hare trappings, as well as mean (`r juvenile_count$mean_count`) and median (`r juvenile_count$median_count`) values with a standard deviation of (`r juvenile_count$sd_count`). There is a notable trend of trapping count observations decline throughout the years. One could conclude a number of different extrapolations from this trend. There should be further investigation moving forward into the possible variables that could impact these count numbers, e.g. the number of days and traps each year, to gain a better perspective from which a more informed analyzation of trends can occur.

#### B. Visual of Juvenile Hare Weights

```{r juvenile_weight_graph, fig.align = "center"}
juvenile %>% 
  ggplot(aes(x = sex, y = weight)) + #Creating graphs to compare weights of juvenile females to males at each grid site.
  geom_beeswarm(aes(color = sex)) +
  facet_wrap("grid") + # Allows you to convey data for more than one observation within a variable
  geom_boxplot(fill = NA, width = 0.2, outlier.color = NA) +
  stat_summary(fun = mean, 
               geom = "point", 
               shape = 20, 
               size = 4, 
               color = "black", 
               fill = "black") +
  labs(x = "Sex",
       y = "Weight(g)",
       title = "Juvenile Snowshoe Hare Weight by Sex at Three Different Sites") +
    scale_x_discrete(labels = c("Female", "Male", "N/A")) +
  theme(legend.position = "none") # take out legend in graph
```

**Figure 2.** Juvenile snowshoe hare weights (grams) compared by sex (Coral (Female), Teal (Male) and Grey (N/A)) across the three different sites data was collected at for study. Weight is depicted on the y axis and visual separation by sex type on the x axis. Results for Third category for sex, N/A, is included.

Figure 3 depicts the differences in weight between sexes at each site location. Visual conveys differing trends at each site. For the bonbs site, female weights vary more significantly with greater distributions of weight than male weights. For the bonmat site, the distribution range of weights between sexes is pretty similar with an exception of slightly lower weight distributions with females. This trend also holds true at the bonrip site, however this site does convey the greatest density of weight distributions between all three sites.  

#### C. Juvenile Weight Comparisons

**Table 2.** Comparing juvenile snowshoe hare weights between male and female sexes with summary statistics that include mean, median, and standard deviation of weights across both sexes, as well as sample size. 
```{r weight_summary_stats}
weight_stats <- juvenile %>% #Creating summary statistics 
  group_by(sex) %>% 
  summarize(mean_weight = round(mean(na.omit(weight)),2), # na.omit takes out N/A observations within a variable
            median_weight = median(na.omit(weight)),
            sd_weight = round(sd(na.omit(weight)), 2),
            n = n())

weight_stats %>% 
  kable(col.names = c("Sex", "Mean weight", "Median weight", "Standard deviation", "Sample Size")) %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r summary_stats_pt3, include=FALSE}
#Create female and male stats to perform exploratory graphs and analyses
snwhares_f <- juvenile %>% 
  filter(sex == "f") %>% 
  pull(weight) %>% 
  na.omit()

snwhares_m <- juvenile %>% 
  filter(sex == "m") %>% 
  pull(weight) %>% 
  na.omit()

# Plot histograms & qq-plots to check assumptions
hist(snwhares_f) # Normal Distribution
qqnorm(snwhares_f) # Linear

hist(snwhares_m) # Somewhat Normal Distribution
qqnorm(snwhares_m) # Linear

#T-test
snwhare_ttest <- t.test(snwhares_f, snwhares_m)

#Cohen's d
snwhare_effsize <- cohen.d(snwhares_f, snwhares_m)

#Get difference in actual means
diff_means <- abs(weight_stats %>% 
  filter(sex == "f") %>% 
  select(mean_weight)) -
  (weight_stats %>% 
  filter(sex == "m") %>% 
  select(mean_weight))
```

Based on mean weight values in grams, observed female juvenile snowshoe hare weights (`r weight_stats[1,2]`) are less than that of observed male weights (`r weight_stats[2,2]`). The difference in means between male and female hares is as follows, (`r abs(diff_means)`) in (g). Cohen's D effect size is very low (`r snwhare_effsize $estimate`), and a t-test gives a p-value of (`r snwhare_ttest$p.value`). 

#### D. Relationship Between Juvenile Weight & Hind Foot Length

```{r hindft_weight_summary_stats, fig.align='center'}
#Linear regression
hndft_weight_lm  <- lm(weight~hindft, data = juvenile) 

#Get tidy versions of the model output to call later on in text
hndft_weight_lm_glance <- glance(hndft_weight_lm) 
hndft_weight_lm_tidy <- tidy(hndft_weight_lm)

# See model diagnostics:
# plot(hndft_weight_lm)

# Get Pearson's r correlation
hndft_weight_cor <- cor.test(juvenile$weight, juvenile$hindft)

# Tidy version of correlation output:
hndft_weight_cor_tidy <- tidy(hndft_weight_cor)

# Graph linear model within a jitter plot
juvenile %>% 
  ggplot(aes(x = hindft, y = weight)) +
  geom_jitter(color = "black") +
   geom_smooth(method = "lm",
              color = "red",
              size = 0.8,
              fill = "orange",
              alpha = 0.2) +
  labs(x = "Hind Foot Length (mm)",
       y = "Weight (g)",
       title = "Relationship between Hind Foot Length (mm) to Weight(g)") +
  theme_minimal() +
  ggpubr::stat_regline_equation(label.x = 75, label.y = 1200)
```

**Figure 3.** Relationship between juvenile snowshoe hare hind foot length (mm) and weight (g). Points indicate individual juvenile snowshoe hare measurements. Linear model summary: $\beta$~1~ = `r round(hndft_weight_lm_tidy$estimate[2], 2)` g mm^-1^, p< .001,R^2^ = `r round(hndft_weight_lm_glance$r.squared,2)`, Pearson's *r* = `r round(hndft_weight_cor_tidy$estimate,2)`.
Data: Kielland et. al. (2017).

The relationship between juvenile snowshoe hare hind foot length (mm) and weight (g) appears relatively linear (Figure 3). Simple linear regression revealed that juvenile snowshoe hare hind foot length does not significantly predict juvenile snowshoe hare body weight. (p < 0.001, R^2^ = `r round(hndft_weight_lm_glance$r.squared,2)`) with an average slope of $\beta$ = `r round(hndft_weight_lm_tidy$estimate[2], 2)` g mm^-1^ (i.e., for each one millimeter increase in hind foot length we expect an average increase in body mass of `r round(hndft_weight_lm_tidy$estimate[2], 2)` g). Hind foot length and body weight are moderately, and significantly, positively correlated (Pearson's *r* = `r round(hndft_weight_cor_tidy$estimate,2)`, p < 0.001). Diagnostic plots (not included) reveal normally distributed and heteroscedastic residuals. 

## Summary

Exploratory data analysis reveals the following initial findings: 

- Female weights vary more significantly with greater distributions of weight than male weights across all three sites.
- Based on mean weight values in grams, observed juvenile female hare weights are less than that of observed juvenile male weights (the difference is significant, and the effect size is small)
- Hind foot length appears linearly related with body weight; the relationship (by simple linear regression) is moderate, with a slope (expected average increase in body weight with a 1 mm increase in hind foot length) of `r round(hndft_weight_lm_tidy$estimate[2], 2)` g mm^-1^.

## Citations

Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. 

